---
title: 'Midterm Project: County-level oil and gas production'
author: "Zichun Liu"
date: "10/11/2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(jsonlite)
library(knitr)
library(ggplot2)
library(reshape2)
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
```


## Part One: Data Description

### 1.1 Background and Data sources

US has large quantities of oil and gas trapped in shale and other formations. 

Prior to the application of unconventional drilling methods, U.S. natural gas production had been growing slowly, and U.S. oil production had been in decline since the mid-1980s. Both trends reversed during the 2000s.

Most policies regarding oil and gas development occur at the State level. Using data from State agencies, the Energy Information Administration (EIA) publishes oil and gas production totals by State, but more local data for the entire United States had not been widely available.

By acquiring disaggregated oil and gas production data from State agencies, a national county-level database was created, providing yearly estimates of onshore production for counties in the lower 48 States.  Nationwide county-level data permit a more comprehensive assessment of the geography of oil and gas development. These county-level data also allow researchers to assess changes in rural production over the past decade.


### 1.2 Description of Dataset

County-level data from oil and/or natural gas producing States—for onshore production in the lower 48 States only—are compiled on a State-by-State basis. 

Most States have production statistics available by county, field, or well, and these data were compiled at the county level to create a database of county-level production, annually for 2000 through 2011.

Raw data for natural gas is for gross withdrawals, and oil data almost always include natural gas liquids. 

In the data file, counties with increases or decreases in excess of $20 million in oil and/or natural gas production during 2000-11 are also identified. See Documentation and Maps for more details.


### 1.3 How it was collected, assembled and maintained

ERS researchers created this national county-level database providing yearly estimates of onshore production for counties in the lower 48 States. And was maintained by ERS researchers as well.

## Part Two: Data Cleaning and Preparation

### 2.1 Manipulate oilgas
```{r}
oilgas <- read.csv2("oilgascounty.csv",header = TRUE,sep = ",")
oilgas$County_Name <- str_replace_all(oilgas$County_Name, " County","")
for (i in 1:length(letters)) {
  oilgas$County_Name <- str_replace_all(oilgas$County_Name, 
                                        LETTERS[i],letters[i])}
for (i in 1:length(state.abb)) {
  oilgas$Stabr <- str_replace_all(oilgas$Stabr, 
                                  state.abb[i],state.name[i])}
for (i in 1:length(letters)) {
  oilgas$Stabr <- str_replace_all(oilgas$Stabr, 
                                  LETTERS[i],letters[i])}
oilgas$oil_change_group <- factor(oilgas$oil_change_group, 
                                  levels = c("H_Growth", "H_Decline", "Status Quo"))
oilgas$gas_change_group <- factor(oilgas$gas_change_group, 
                                  levels = c("H_Growth", "H_Decline", "Status Quo"))
oilgas$oil_gas_change_group <- factor(oilgas$oil_gas_change_group, 
                                      levels = c("H_Growth", "H_Decline", "Status Quo"))

head(oilgas)
#see if id is unique
length(unique(oilgas$geoid)) == length(oilgas$geoid)
#see if FIPS is same as geoid
sum(oilgas$FIPS - oilgas$geoid)
#remove FIPS
oilgas$FIPS <- NULL

#calculate total withdraw in 12 years and level the sum
for (i in 1:length(oilgas$geoid)) {
  oilgas$ttoil[i] <- sum(as.numeric(oilgas[i,8:19]))
  a <-oilgas$ttoil[i]
  if (a < 10) {oilgas$oil_level[i] <- (0)}
  else if (a >= 10 & a < 1000) {oilgas$oil_level[i] <- (1)}
  else if (a >= 1000 & a < 10000) {oilgas$oil_level[i] <- (2)}
  else if (a >= 10^4 & a < 10^5) {oilgas$oil_level[i] <- (3)}
  else if (a >= 10^5 & a < 10^6) {oilgas$oil_level[i] <- (4)}
  else if (a >= 10^6 & a < 10^7) {oilgas$oil_level[i] <- (5)}
  else if (a >= 10^7 & a < 10^8) {oilgas$oil_level[i] <- (6)}
  else if (a >= 10^8 & a < 10^9) {oilgas$oil_level[i] <- (7)}
  else {oilgas$oil_level[i] <- (8)}
  oilgas$ttgas[i] <- sum(as.numeric(oilgas[i,20:31]))
  a <- oilgas$ttgas[i]
  if (a < 10 | is.na(a) == TRUE) {oilgas$gas_level[i] <- (0)}
  else if (a >= 10 & a < 1000) {oilgas$gas_level[i] <- (1)}
  else if (a >= 1000 & a < 10000) {oilgas$gas_level[i] <- (2)}
  else if (a >= 10^4 & a < 10^5) {oilgas$gas_level[i] <- (3)}
  else if (a >= 10^5 & a < 10^6) {oilgas$gas_level[i] <- (4)}
  else if (a >= 10^6 & a < 10^7) {oilgas$gas_level[i] <- (5)}
  else if (a >= 10^7 & a < 10^8) {oilgas$gas_level[i] <- (6)}
  else if (a >= 10^8 & a < 10^9) {oilgas$gas_level[i] <- (7)}
  else if (a >= 10^9){oilgas$gas_level[i] <- (8)}
}

```

### 2.2 Get first tidy dataset
```{r}
a <- oilgas %>% gather(key = year,value = oil, oil2000:oil2011)
b <- oilgas %>% gather(key = year,value = gas, gas2000:gas2011)
b <- b[c("geoid","year","gas")]
target <- cbind(a,b[-1])
target <- target[-c(8:19,29)]
target$year <- factor(str_replace_all(target$year, "oil",""))
head(target)
saveRDS(target,"tidydataoilgas.rds")
rm(a,b,target)
tidy <- readRDS("tidydataoilgas.rds")
```

### 2.3 Get another tidy dataset
```{r}
tidy2 <- oilgas[,-(8:31)]
saveRDS(tidy2,"data_without_year.rds")
```

## Part Three: Exploratory Data Analysis
### 3.1 Explore on first tidy dataset(oil&gas data from 2000 to 2011)
#### 1) on nation level
```{r warning=FALSE, fig.align='left', fig.height=3, fig.width=10}
#on nation level
Nationoil <- NULL
inrateoil <- NULL
Nationgas <- NULL
inrategas <- NULL
for (i in 1:12) {
  Nationoil[i] <- sum(as.numeric(subset(tidy, subset = year == 1999+i)$oil))
  Nationgas[i] <- sum(as.numeric(subset(tidy, subset = year == 1999+i)$gas))
  if (i == 1) {
    inrateoil[i] <- 0
    inrategas[i] <- 0
  }
  else{
    inrateoil[i] <- (Nationoil[i]-Nationoil[i-1])/Nationoil[i-1]
    inrategas[i] <- (Nationgas[i]-Nationgas[i-1])/Nationgas[i-1]
  }
}
Nation <- data.frame(year=2000:2011, 
                     Nationoil = Nationoil,
                     oilinrate = inrateoil, 
                     Nationgas = Nationgas, 
                     gasinrate = inrategas)

rm(Nationoil,Nationgas,inrategas,inrateoil)

summary(Nation)

#average increasing rate of oil production
sum(Nation$oilinrate)/11
#average increasing rate of gas production
sum(Nation$gasinrate)/11

#nation <- Nation %>% gather(key = resource, value = volume, c(Nationoil,Nationgas))

ggplot(data = Nation, aes(x = year, y = Nationoil)) + 
  geom_point(aes(color = factor(year))) + 
  geom_smooth(linetype = 2, size = 0.5, alpha = 0)
ggplot(data = Nation, aes(x = year, y = oilinrate)) + 
  geom_bar(stat = "identity", aes(fill = factor(year))) + 
  geom_smooth(linetype = 2, size = 0.5, alpha = 0)

ggplot(data = Nation) + 
  geom_point(aes(x = year, y = Nationgas, color = factor(year))) + 
  geom_smooth(aes(x = year, y = Nationgas),linetype = 2, size = 0.5, alpha = 0)
ggplot(data = Nation, aes(x = year, y = gasinrate)) + 
  geom_bar(stat = "identity", aes(fill = factor(year))) + 
  geom_smooth(linetype = 2, size = 0.5, alpha = 0)
```

#### 2) overall
```{r warning=FALSE, fig.align='left', fig.height=3, fig.width=10}
#overall
ggplot(tidy) + 
  geom_point(aes(x = year, y = oil, color = oil_change_group))
ggplot(tidy) + 
  geom_point(aes(x = year, y = gas, color = gas_change_group))
```

#### 3) on state level(use example of Wyoming)
```{r warning=FALSE, fig.align='left', fig.height=3, fig.width=10}
#by state(use example of Wyoming)
ggplot(data = subset(tidy,subset = Stabr=="wyoming"))+ 
  geom_point(aes(x = year, y = oil, color = oil_change_group))
ggplot(data = subset(tidy,subset = Stabr=="wyoming"))+ 
  geom_point(aes(x = year, y = gas, color = gas_change_group))
```

#### 4) on County_Name (use example of baldwin, Alabama)
```{r warning=FALSE, fig.align='left', fig.height=3, fig.width=10}
ggplot(data = subset(tidy,subset = geoid==1003), 
       aes(x = as.integer(year), y = oil))+ 
  geom_point(aes(color = factor(year))) + 
  geom_smooth(linetype = 2, size = 0.5, alpha = 0)
ggplot(data = subset(tidy,subset = geoid==1003), 
       aes(x = as.integer(year), y = gas)) + 
  geom_point(aes(color = factor(year))) + 
  geom_smooth(linetype = 2, size = 0.5, alpha = 0)

```

#### 5) on year level(use example of first year(2000) and last year(2011))
```{r warning=FALSE, fig.align='left', fig.height=3, fig.width=10}
#oid
ggplot(data = subset(tidy,subset = (year == 2000)), 
       aes(x = oil_change_group, y = oil, 
           color = Rural_Urban_Continuum_Code_2013)) + 
  geom_jitter()
ggplot(data = subset(tidy,subset = (year == 2011)), 
       aes(x = oil_change_group, y = oil, 
           color = Rural_Urban_Continuum_Code_2013)) + 
  geom_jitter()

#gas
ggplot(data = subset(tidy,subset = (year == 2000)), 
       aes(x = gas_change_group, y = gas, 
           color = Rural_Urban_Continuum_Code_2013)) + 
  geom_jitter()
ggplot(data = subset(tidy,subset = (year == 2011)), 
       aes(x = gas_change_group, y = gas, 
           color = Rural_Urban_Continuum_Code_2013)) + 
  geom_jitter()
```

#### 6) on change_group
```{r warning=FALSE, fig.align='left', fig.height=3, fig.width=10}
#oil 
ggplot(data = subset(tidy,subset = oil_change_group=="H_Growth"))+ 
  geom_point(aes(x = year, y = oil, 
                 color = Rural_Urban_Continuum_Code_2013))
ggplot(data = subset(tidy,subset = oil_change_group=="H_Decline"))+
  geom_point(aes(x = year, y = oil, 
                 color = Rural_Urban_Continuum_Code_2013))
ggplot(data = subset(tidy,subset = oil_change_group=="Status Quo"))+ 
  geom_point(aes(x = year, y = oil, 
                 color = Rural_Urban_Continuum_Code_2013))

#gas
ggplot(data = subset(tidy,subset = gas_change_group=="H_Growth"))+ 
  geom_point(aes(x = year, y = gas, 
                 color = Rural_Urban_Continuum_Code_2013))
ggplot(data = subset(tidy,subset = gas_change_group=="H_Decline"))+ 
  geom_point(aes(x = year, y = gas, 
                 color = Rural_Urban_Continuum_Code_2013))
ggplot(data = subset(tidy,subset = gas_change_group=="Status Quo"))+ 
  geom_point(aes(x = year, y = gas, 
                 color = Rural_Urban_Continuum_Code_2013))

```

### 3.2 Explore on the second tidy dataset(non-year variables)
#### 1) State Status
```{r}
ggplot(data = tidy2, aes(x = oil_gas_change_group, fill = Stabr)) + 
  geom_bar(position = "dodge")
```

#### 2) Choropleth map
```{r warning=FALSE}
#pullout the level and region information
pullout <- oilgas[,c(2,3,36,38)]
colnames(pullout)[1] <- "region"
colnames(pullout)[2] <- "subregion"

#get map data for US counties and states
county_map <- map_data("county")
state_map <- map_data("state")

#merge pullout and county_map
pullout_map <- merge(county_map, pullout, 
                     by.x=c("region", "subregion"), 
                     by.y=c("region", "subregion"),  all.x=TRUE)

#resort merged data
pullout_map <- arrange(pullout_map, group, order)

#relpace NA with 0's
pullout_map[is.na(pullout_map)] <- 0

#generate a disctrete color pallette    
pal_red <- c("#ffffff","#fbb5bf","#f88b9e","#f67192",
             "#f13367","#f23b7a","#eb004e","#ea0039","#cb0036")
pal_blue <- c("#ffffff","#bfdeec","#9ac4e0","#3778b4",
              "#233482","#222e70","#15275b","#021430","#000000")

theme_clean <- function(base_size = 12) {
  require(grid)
  theme_grey(base_size) %+replace%
    theme(
      axis.title      =   element_blank(),
      axis.text       =   element_blank(),
      panel.background    =   element_blank(),
      panel.grid      =   element_blank(),
      axis.ticks.length   =   unit(0,"cm"),
      axis.ticks.margin   =   unit(0,"cm"),
      panel.margin    =   unit(0,"lines"),
      plot.margin     =   unit(c(0,0,0,0),"lines"),
      complete = TRUE
    )
}

#choropleth map on oil level
ggplot( pullout_map, aes( x = long , y = lat , group=group ) ) +
  geom_polygon(linetype = 2, size = 0.1, colour = "lightgrey",
               aes( fill = factor(oil_level) ) ) +
  scale_fill_manual( values = pal_red ) +
  expand_limits( x = pullout_map$long, y = pullout_map$lat ) +
  coord_map( "polyconic" ) + 
  labs(fill="Total Oil Production") + 
  theme_clean( )+
  geom_path( data = state_map, color = "darkgrey")


#choropleth map on gas level
ggplot( pullout_map, aes( x = long , y = lat , group=group ) ) +
  geom_polygon(linetype = 2, size = 0.1, colour = "lightgrey" , 
               aes( fill = factor(gas_level) ) ) +
  scale_fill_manual( values = pal_blue ) +
  expand_limits( x = pullout_map$long, y = pullout_map$lat ) +
  coord_map( "polyconic" ) + 
  labs(fill="Total Gas Production") + 
  theme_clean( )+
  geom_path( data = state_map, color = "darkgrey")

rm(a,i)
```


